---
title: "Decision Trees"
author: "INSOFE Lab Activity on Decision Trees"
date: "23 July 2017"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# C5.0 Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}
rm(list = ls(all = TRUE))

```

## Goal

* The goal of this activity is to predict wether a patient has liver disease or not based on various patient related attributes


## Agenda 

* Get the data

* Data Pre-processing

* Build a model

* Predictions

* Communication


## Reading & Understanding the Data

### Read the Data

Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}
setwd("I:/DATA-SCIENCE/Insofe/Assignments/DecisionTree")
patient.data = read.csv("ilpd_data.csv", na.strings = c(NA,"") )

```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 582 observations and 11 variables

- __The variable descriptions are given below:__

1 - age	 : Age of the patient

2 - gender : Gender of the patient

3 - TB : Total Bilirubin content

4 - DB : Direct Bilirubin content

5 - alk_phos : Alkaline Phosphotase content

6 - alamine : Alamine Aminotransferase content

7 - aspartate : Aspartate Aminotransferase content

8 - TP : Total Protiens content

9 - albumin : 	Albumin content

10 - A/G : Ratio of Albumin and Globulin 

11 - Disease : Whether the patient has liver disease or not 


```{r}
str(patient.data)
summary(patient.data)
```

```{r}

head(patient.data)
tail(patient.data)
```


## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}
sum(is.na(patient.data))

```

* Verify the data types assigned to the variables in the dataset

```{r}

str(patient.data)

```

### Split the Data into train and test sets

* Use stratified sampling to split the data into train/test sets (70/30)

* Use the createDataPartition() function from the caret package to do stratified sampling

```{r}
library(caret)
train.rows = createDataPartition(patient.data$disease, p = 0.7, list = FALSE)
train.data = patient.data[train.rows,]
test.data = patient.data[-train.rows,]
```

### Impute the missing values

* Impute missing values using knnImputation() function in both the train and test datasets

```{r}
library(DMwR)
train.data = knnImputation(train.data, k = 7, scale = T)
test.data = knnImputation(test.data, k =7, scale = T, distData = train.data)

```

## Build a  Decision Tree

### Model the tree

* Use Quinlan's C5.0 decision tree algorithm implementation from the C50 package to build your decision tree

```{r}
library(C50)
c5_tree = C5.0(disease ~ ., train.data )

```

* Build a rules based tree

```{r}
c5_rules = C5.0(disease ~ ., train.data, rules = TRUE)

```

### Variable Importance in trees

* Find the importance of each variable in the dataset

```{r}
C5imp(c5_tree, metric = "usage")

```

### Rules from trees

* Understand the summary of the returned c5.0 rules based on the decision tree model


```{r}
summary(c5_rules)

```


### Plotting the tree

* Call the plot function on the tree object to visualize the tree

```{r, fig.width= 35, fig.height=15}

plot(c5_tree)

```


## Evaluating the model

### Predictions on the test data

* Evaluate the decision tree using the standard error metrics on test data

```{r}
preds = predict(c5_tree, test.data)


```

* Report error metrics for classification on test data

```{r}

library(caret)
confusionMatrix(preds, test.data$disease)

```

# CART Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}

rm(list = ls(all = TRUE))

```

* The classification and regression trees use gini index in place of the gain ratio (based on information gain) used by the ID3 based algorithms, such as c4.5 and c5.0

## Goal

* The goal of this activity is to predict the heating load of a residential building, if the building parameters are given

* Hence, in the future architects would be able to build more energy efficient buildings as they can optimize the building parameters to reduce the heating load

## Agenda 

* Get the data

* Data Pre-processing

* Build a model

* Predictions

* Communication


## Reading & Understanding the Data

### Read the Data

* Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}
setwd("I:/DATA-SCIENCE/Insofe/Assignments/DecisionTree")
building.data = read.csv("building_energy.csv")

```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 768 observations and 9 variables

```{r}
str(building.data)
summary(building.data)
head(building.data)
tail(building.data)

```

* The variable names are self explanatory, for further information visit http://www.sciencedirect.com/science/article/pii/S037877881200151X

## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}
sum(is.na(building.data))

```

* Verify the data types assigned to the variables in the dataset

```{r}

# Enter answer here
str(building.data)

```

### Split the Data

* Split the data into train/test sets (70/30)

```{r}
library(caret)
train.rows = createDataPartition(building.data$heating_load, p = 0.7, list = FALSE)
train.data = building.data[train.rows,]
test.data = building.data[-train.rows,]


```

## Build a Regression Tree

### Model the tree

* Use the rpart package to build a cart tree to predict the heating load

```{r}

library(rpart)
reg.tree = rpart(heating_load ~ ., train.data)
printcp(reg.tree)

```

### Tree Explicability

* Print the variable importance

```{r}
reg.tree$variable.importance


```

* Plot the regression tree

```{r, fig.width=8, fig.height=5}

library(rpart.plot)
rpart.plot(reg.tree)

```

## Evaluation on Test Data

* Report error metrics on the test data

```{r}

preds.reg = predict(reg.tree, test.data)

```

```{r}
library(DMwR)
regr.eval(test.data$heating_load, preds.reg)


```




















